---
output:
  html_document: default
  pdf_document: default
---
Loan Data Analysis by Jed
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}


library(dplyr)
library(ggplot2)
library(knitr)
library(GGally)
library(scales)
knitr::opts_chunk$set(fig.width=9,fig.height=6,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('prosperLoanData.csv')
df<- select(df,Term,BorrowerAPR, LenderYield, EstimatedReturn, ProsperRating..Alpha., IsBorrowerHomeowner,CreditScoreRangeLower,TotalCreditLinespast7years,DelinquenciesLast7Years,IncomeVerifiable,IncomeRange,Recommendations,EstimatedEffectiveYield,Investors,StatedMonthlyIncome,TradesNeverDelinquent..percentage.,LoanOriginalAmount,LP_ServiceFees)

```


# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots}
#Load basic stats about the data and plot a histogram of credit score
str(df)

ggplot(data = subset(df,df$CreditScoreRangeLower>0),
       aes(x = CreditScoreRangeLower))+
  geom_histogram(binwidth = 20)

```

This looks like a very normal distribution. There is a slight jump at 600 
perhaps because it qualifies them for certain benefits but I am not 100% sure. 
Most of the credit scores lie in the range of 620 - 750. There also seem to be 
a few credit scores over 850 which is supposed to be the maximum.

```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Effective Yield histogram
ggplot(data = df,
       aes(x=EstimatedEffectiveYield))+
  geom_histogram(binwidth = .05)
```

Estimated effective yield is the amount of money the lender hopes to make from
the loan. No one wants to loan money to somone that is going to have a negative yield so 
this makes sense.


```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Reorder the Prosper ratings and plot

df$ProsperRating..Alpha. <- ordered(df$ProsperRating..Alpha. , levels = c('','AA','A','B','C','D','E','HR'))
ggplot(data = subset(df,ProsperRating..Alpha.!=''),
       aes(x=ProsperRating..Alpha.))+
  geom_histogram(stat = 'count')
```

I had to get rid of all of the listings without a Prosper number and then 
resort with AA as the best rating. This distribution looks about right with AA 
being the least common. Maybe people with AA ratings don't need to go to 
Prosper?

```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Investors Histogram
ggplot(data = df,
       aes(x=Investors))+
  geom_histogram(binwidth = 1)+
  coord_cartesian(xlim = c(0,100))
```

Number of investors for each loan. Almost all people that needed loans only needed 1 investor.


```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Make sure that incomes are verified before plotting

ggplot(data = subset(df,StatedMonthlyIncome>0),
       aes(x=StatedMonthlyIncome))+
  geom_histogram(binwidth = 500)+
  coord_cartesian(xlim= c(0,quantile(df$StatedMonthlyIncome,.99)))


summary(df$StatedMonthlyIncome)
```
This is the distribution of borrowers incomes. They are not all verified so I will sort
by verified incomes in the below plot.

```{r echo=FALSE,message=FALSE, warning=FALSE,}



ggplot(data = subset(df,StatedMonthlyIncome>0 & IncomeVerifiable =="True"),
       aes(x=StatedMonthlyIncome))+
  geom_histogram(binwidth = 500)+
  coord_cartesian(xlim= c(0,quantile(df$StatedMonthlyIncome,.99)))
  
summary(subset(df,StatedMonthlyIncome>0 & IncomeVerifiable =="True")$StatedMonthlyIncome)
```

The median Stated monthly income was 4667. The extremely rich have stated 
monthly incomes over 1 million so a mean would be a better measure. After 
subseting by people who have verified incomes, we see that the max has gone 
down but the mean and median are about the same.




```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Difference between homeowners?
ggplot(data = df,
       aes(x=Term))+
  geom_bar()+
  scale_x_continuous(breaks = c(12,36,60))
```
The terms of all loans. Barely any seem to be 12 months for some reason.

```{r echo=FALSE,message=FALSE, warning=FALSE,}

ggplot(data = subset(df,IsBorrowerHomeowner== 'True'),
       aes(x=Term))+
  geom_bar()+
  scale_x_continuous(breaks = c(12,36,60))
```

I was wondering if people who already owned homes changed had different ideas 
on how long they needed a term to be. It doesn't look like there is any 
difference between homeowners and non-homeowners.

```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Sort by different Prosper ratings to see differences

ggplot(data = df,
       aes(x=BorrowerAPR))+
  geom_histogram(binwidth = .01)
```
I think APR will be a main focus on my analysis. APR is a good representation 
of the risk each loan has. If a loan is risk and might not be paid back, a high 
APR will be necessary. Just looking at the APR distribution without subsetting, 
the distribution looks normal.

```{r echo=FALSE,message=FALSE, warning=FALSE,}

ggplot(data = subset(df,ProsperRating..Alpha. == 'AA'),
       aes(x=BorrowerAPR))+
  geom_histogram(binwidth = .01)
```
However, looking at the data after subsetting by Prosper rating changes the story.

```{r echo=FALSE,message=FALSE, warning=FALSE,}

ggplot(data = subset(df,ProsperRating..Alpha. == 'E'),
       aes(x=BorrowerAPR))+
  geom_histogram(binwidth = .01)
```
Higher Prosper rating clearly leads to lower APR.

```{r echo=FALSE,message=FALSE, warning=FALSE,}

ggplot(data =  subset(df,IsBorrowerHomeowner== 'True'),
       aes(x=BorrowerAPR))+
  geom_histogram(binwidth = .01)


```

When we subset by different Prosper ratings, we see that higher ratings lead to lower APR. There is also a spike at 
36% APR for some reason. I will be looking into this variable more in the next 
analysis. The homeowner subset looks very similar to our original plot though.

# Univariate Analysis

### What is the structure of your dataset?

Out of all of the variables we have to explore, I will mainly be focusing on
the following:
Term (12,36,60 months)
BorrowerAPR
LenderYield
EstimatedReturn
ProsperRating (AA, A,B ,C ,D ,E, HR from best -> worst)
IsBorrowerHomeowner (T/F)
CreditScoreRangeLower
TotalCreditLinespast7years
DelinquenciesLast7Years
IncomeVerifiable (T/F)
IncomeRange
Recommendations

Most loans are 36 month loans and very few loans are 12 months.

Almost everyone has their income verified

About half the loans are made by homeowners.

### What is/are the main feature(s) of interest in your dataset?

The main features of this dataset are the APR and how each variable will affect
the APR.

### What other features in the dataset do you think will help support your 
investigation into your feature(s) of interest?



### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? 
Did you perform any operations on the data to tidy, adjust, or change the form 
of the data? If so, why did you do this?


# Bivariate Plots Section


```{r echo=FALSE,message=FALSE, warning=FALSE,fig.width=10, fig.height=10, Bivariate_Plots}
theme_set(theme_minimal(10))

set.seed(42)

df_subset <- df[c(1:4,7:10,12)]#add more variables you want to measure here
ggpairs(df_subset[sample.int(nrow(df_subset),100),])+
  theme(text = element_text(size= 8))

```
Paired data to see a broad view of relationships between variables.

```{r echo=FALSE,message=FALSE, warning=FALSE}
#APR vs Credit score 

ggplot(data = df,
       aes(x = BorrowerAPR, y = CreditScoreRangeLower))+
  geom_point()
```

There are a lot of values that are 0 that are probably not valid data points. 
I will remove those as well as lowering the alpha of the point to see where 
more of the points are distributed.


```{r echo=FALSE,message=FALSE, warning=FALSE}
#Add a smoother

ggplot(data = subset(df,CreditScoreRangeLower>0),
       aes(x = BorrowerAPR, y = CreditScoreRangeLower))+
  geom_jitter(alpha = .03)+
  geom_smooth(color = 'light Blue')
```
It looks like the smoother is fairly consistent from the 5% APR range to around the 25%.
There seems to be a negative correlation between APR and credit score.

```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Statistical analysis
with(subset(df,CreditScoreRangeLower>0),cor.test(BorrowerAPR,CreditScoreRangeLower))
with(subset(df,CreditScoreRangeLower>0 & BorrowerAPR< .25 & BorrowerAPR >.05),cor.test(BorrowerAPR,CreditScoreRangeLower))

summary(subset(select(df,BorrowerAPR,CreditScoreRangeLower),CreditScoreRangeLower>0))
   
```

Here I have subset the data and changed the alpha to .03 since we have so many 
datapoints. It looks like we can see that there is some significance in the 
data. The main thing I see is that the region with APR between 5% and 25% seems 
to have more correlation with credit score than the region above 25%.


```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Different boxplots to see Prosper rating effects

ggplot(data = subset(df,ProsperRating..Alpha.!= ''),
       aes(x= ProsperRating..Alpha., y=BorrowerAPR))+
  geom_boxplot()
```
Borrower APR boxplot sorted by Prosper rating. There is strong correlation.

```{r echo=FALSE,message=FALSE, warning=FALSE,}

ggplot(data = subset(df,ProsperRating..Alpha.!= ''),
       aes(x= ProsperRating..Alpha., y=LenderYield))+
  geom_boxplot()


subset(df,ProsperRating..Alpha.!= '')%>%
  group_by(ProsperRating..Alpha.)%>%
  summarise(mean_APR = mean(BorrowerAPR),
            mean_yield = mean(LenderYield),
            n = n())

```

As we can see with these boxplots, the prosper rating has a very clear effect 
on what the APR rate will be. The APR and Lender yield will obviously have very 
similar values. APR will always be higher than the yield though, just because 
Prosper has to take a cut.


```{r echo=FALSE,message=FALSE, warning=FALSE}
#Change the order of IncomeRange and make a boxplot

df$IncomeRange <- ordered(df$IncomeRange , levels = c('Not employed','$0','$1-24,999','$25,000-49,999','$50,000-74,999','$75,000-99,999','$100,000+','Not displayed'))

ggplot(data = subset(df,IncomeRange!= 'Not displayed' & IncomeVerifiable == "True"),
       aes(x= IncomeRange, y=BorrowerAPR))+
  geom_boxplot()+
  theme(text = element_text(size= 10))

table(subset(df,IncomeRange!= 'Not displayed' & IncomeVerifiable == "True")$IncomeRange)
```

I had initially made the error of not subsetting the data by only incomes that 
had been verified which would probably have had incorrect data. Now this plot 
is interesting because for some reason, the unemployed people have a very low 
APR. This doesn't seem right but after looking at the table, we can see that 
there are so few people that are unemployed or with $0 of net income, that it 
would probably make the data inaccurate. Everything else in the graph seems 
very reasonable and it makes sense that people with higher incomes would be 
able to secure loans with less APR.

```{r echo=FALSE,message=FALSE, warning=FALSE}
#Make sure to set alpha lower to see density if different graphs

ggplot(data = df,
       aes(x= IsBorrowerHomeowner, y=BorrowerAPR))+
  geom_boxplot()
```

I plotted some other insights from the data but these do not seem as conclusive
as the previous plots. Homeowners do have less APR on average but only by
around 4%.

```{r echo=FALSE,message=FALSE, warning=FALSE}


ggplot(data = subset(df,DelinquenciesLast7Years<50),
       aes(x= DelinquenciesLast7Years, y= BorrowerAPR))+
  geom_point(alpha = .05)
```
The delinquencies of the last 7 years do not seem to have a huge impact on the 
APR but most people have 0. Maybe this info is not very visible to the lender? 


```{r echo=FALSE,message=FALSE, warning=FALSE}

ggplot(data = df,
       aes(x= BorrowerAPR, y= LoanOriginalAmount))+
  geom_point(alpha = .05)



```


Most loan amounts are less than 10k and those loans seem to vary in APR. This 
may mean that people do not want to lend too much money to those with high 
risk. Why do all of the 35k loan amounts have similar APR's? It is interesting 
to see that as the amount of the loan increases, the average APR range 
decreases. Maybe only financially stable people borrow large amounts of money?



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
investigation. How did the feature(s) of interest vary with other features in 
the dataset?

My main focus of this dataset has been how the interest a borrower has to pay 
is affected by their various characteristics. Many of my initial thoughts did 
end up having an affect such as having a higher income or credit score. 
Some things such as the number of delinquencies and being a home owner has 
less of an affect that I would have thought. 
  
  ### Did you observe any interesting relationships between the other features 
  (not the main feature(s) of interest)?
  
  I thought I would have been able to find a formula for the service fee 
  fairly easily but the service fee might be a more complicated formula.
  
  ### What was the strongest relationship you found?
  
  The strongest relationship was the one between APR and and the Prosper 
  rating. I think this would probably be because Prosper would advertise their 
  own rating much harder than they would any other metric. People also might 
  not want to look too deeply into other metrics on their own.
  
# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
#Clearly show how income range affects both APR and Credit scores

ggplot(data = subset(df,IncomeRange!= 'Not displayed'),
       aes(x= BorrowerAPR, y= CreditScoreRangeLower))+
  geom_point(aes(color = IncomeRange), alpha = .1, size = 2)+
  theme(text = element_text(size= 13))+
  scale_color_brewer(type = "div", palette = "RdYlBu") +
  guides(color=guide_legend(title="Income Range", override.aes = list(alpha = 1)))

```

We can see here that people with higher incomes are also in the higher range 
of the credit score range. This obviously makes sense as the more money you 
have the less likely you are to default on payments. The higher values on the 
APR range also tend to be from people that make less money. This kind of lends 
to the idea that it costs a lot of money to be poor.
```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Change term to a factor to make it easier to plot
  
df$Term <- as.factor(df$Term)
ggplot(data = subset(df,IncomeRange!= 'Not displayed'& StatedMonthlyIncome<250000),
       aes(x= BorrowerAPR, y= StatedMonthlyIncome))+
  geom_point(aes(color = Term),alpha = .1)+
  theme(text = element_text(size= 13))+
  coord_cartesian(ylim = c(0,100000))
```

It looks like people that take out shorter term loans tend to have lower APR. 
Maybe this is because there is less risk in shorter loans? It does not seem 
conclusive though.

```{r echo=FALSE}
#Create a service percentage variable by using service fees / original amount

df$Service_percentage <- -df$LP_ServiceFees/df$LoanOriginalAmount
```
I created a new variable to see the percentage fee that Prosper takes per loan. I 
will look into different possibilities below.

```{r echo=FALSE}

ggplot(data = subset(df,IncomeVerifiable == "True"&StatedMonthlyIncome<50000),
       aes(x= StatedMonthlyIncome, y= Service_percentage))+
  geom_point(alpha=.05)

ggplot(data = df,
       aes(x=LoanOriginalAmount, y=Service_percentage))+
  geom_point(alpha=.05)

ggplot(data = df,
       aes(x=Investors, y=Service_percentage))+
  geom_point(alpha=.05)
```

I wanted to create a variable to see how the service fee was calculated but I
was not able to see a direct correlation with any of the variables. The service
fee is almost always under 2% though.
```{r echo=FALSE}
summary(df$Service_percentage)
```

Summary of Service fee percentage.

```{r echo=FALSE,message=FALSE, warning=FALSE,}
#Facet wrap with Incomerange

ggplot(data = subset(df,IncomeRange != 'Not displayed' & ProsperRating..Alpha.!=''),
       aes(x= ProsperRating..Alpha.,y= BorrowerAPR))+
  geom_boxplot(alpha = .1)+
  facet_wrap(~IncomeRange)
```

The pattern of APR is generally the same no matter your income level. The main 
thing that affects the APR is still the Prosper rating.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the 
investigation. Were there features that strengthened each other in terms of 
looking at your feature(s) of interest?

I think adding a layer showing the income range along with APR and the credit 
score really solidified how higher APR is for the poorer. I also saw that 
the Prosper rating means a lot to how APR is calculated.

### Were there any interesting or surprising interactions between features?

I was surprised that the service fee was not clearly stated or easily 
calculated. I thought Prosper would be upfront about what they charge.


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE,message=FALSE, warning=FALSE, Plot_One}
df$ProsperRating..Alpha. <- ordered(df$ProsperRating..Alpha. , levels = c('','AA','A','B','C','D','E','HR'))
ggplot(data = subset(df,ProsperRating..Alpha.!=''),
       aes(x=ProsperRating..Alpha.))+
  geom_histogram(stat = 'count')+
  ggtitle("Distribution of Proser ratings")
  
  
```

### Description One
The distribution of Prosper ratings is very normal. AA seems to be a bit more 
rare than expected.

### Plot Two
```{r echo=FALSE,message=FALSE, warning=FALSE, Plot_Two}
ggplot(data = subset(df,IncomeRange!= 'Not displayed'),
       aes(x= BorrowerAPR, y= CreditScoreRangeLower))+
  geom_point(aes(color = IncomeRange), alpha = .1)+
  theme(text = element_text(size= 13))+
  ggtitle("APR based on Credit Score and Income Range")+
  scale_x_continuous(labels = percent)
```

### Description Two
This plot seems to suggest that lower income people tend to have lower credit 
scores. There is some slight correlation between credit scores and APR.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = subset(df,ProsperRating..Alpha.!= ''),
       aes(x= ProsperRating..Alpha., y=BorrowerAPR,fill=ProsperRating..Alpha.))+
  geom_boxplot(alpha = .1)+
  theme(text = element_text(size= 13))+
  ggtitle('APR based on Prosper rating')+
  scale_y_continuous(labels = percent)
```

### Description Three
APR seems to be most dependant on the Prosper rating. There are clear cut 
percentages for each rating. 
------
  
  # Reflection

I thought that this dataset was very different than previous datasets. There 
were so many different variables that it made things hard to connect. I chose 
to look at APR because that was what I thought I understood the most. A lot 
of the other data seemed like it needed some research to really understand. 
I think I was able to determine strong relationships between different APR and 
other variables. I'm not sure why the service fee was hard to figure out but 
there must be a lot of variables going into that calculation. In the future, I 
think I could find more relationships between other variables like 
delinquencies and the amount of open lines of credit. It seems like I could find
a lot of credit score and amount of delinquencies. It would be expected that 
higher income leads to less delinquencies. I could also have looked at collection
fees and what imcome/credit group has higher fees.

I think learning and all of these statistical tools was fairly difficult. I think
the idea of plotting was fairly easy but the code to adjust different things was 
harder than expected. I also thought data like this would be more straight 
forward in terms of how easy it is to find trends but it was more difficult than 
I thought. Not everything in the real world is simple so I guess this is why 
companies need data analysts. The biggest thing I would like to improve on is 
creating plots that are fun to look at while being useful at the same time.




  markdown::('R project.R')
